Recovery of a deleted SARS-CoV-2 deep sequencing data set from the early Wuhan epidemic

### 2021-06-15 paper/paper.tex
add trees to results

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/deltadist_RaTG13.pdf}
\caption{The reported collection date of each SARS-CoV-2 sequence versus its relative mutational distance from the RaTG13 bat coronavirus outgroup.
Mutational distances are relative to the putative progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
The plot shows all sequences in GISAID collected no later than February 28, 2020, after performing the quality-control described in the Methods.
Sequences that the joint WHO-China report~\citep{WHO2021origins} describes as being associated with the Wuhan Seafood Market are plotted with squares.
Points are slightly jittered on the y-axis.
Go to {\color{red} add link} for an interactive version of this plot that enables toggling of the outgroup to RpYN06 and RmYN02, mouseovers to see details for each point including strain name and mutations relative to proCoV2, and adjustment of the y-axis jittering.
Static versions of the plot with RpYN06 and RmYN02 outgroups are in Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}.
}
\label{fig:deltadist_RaTG13}
\end{figure*}

@@

This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is chosen to be closest to the bat coronavirus outgroup.
If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three possible ways to root the tree in this fashion (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).

@@

>>>>

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/deltadist_RaTG13.pdf}
\caption{The reported collection date of each SARS-CoV-2 sequence versus its relative mutational distance from the RaTG13 bat coronavirus outgroup.
Mutational distances are relative to the putative progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
The plot shows sequences in GISAID collected no later than February 28, 2020.
Sequences that the joint WHO-China report~\citep{WHO2021origins} describes as being associated with the Wuhan Seafood Market are plotted with squares.
Points are slightly jittered on the y-axis.
Go to {\color{red} add link} for an interactive version of this plot that enables toggling of the outgroup to RpYN06 and RmYN02, mouseovers to see details for each point including strain name and mutations relative to proCoV2, and adjustment of the y-axis jittering.
Static versions of the plot with RpYN06 and RmYN02 outgroups are in Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}.
}
\label{fig:deltadist_RaTG13}
\end{figure*}

@@

@@

This conundrum can also be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is closest to the bat coronavirus outgroup.
If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the bat coronavirus outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the three rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with inferences based on the dates of sequence collection and the epidemiological data suggesting Wuhan was the origin of the pandemic~\citep{pipes2021assessing}.

Tables 6 and 7 of WHO-China report indicate viruses from Huanan Market all in clade B except S13, which has T8782, T28144, and some mutation at 4946. However, clades are connected by single mutations, for instance Sichuan/SC-MY-054/2020 has T8782 but C28144.
